---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains example use cases for `taxonbridge`:

## Table of contents

**1. Detecting ambiguity and inconsistencies**

**2. Annotating a custom taxonomy**

**3. Visualizing a custom taxonomy**

## 1. Detecting ambiguity and inconsistencies

The first example illustrates how to detect ambiguity and inconsistency in a merged taxonomy. Start by loading the 2000 row sample dataset that comes with `taxonbridge`:

```{r, include=TRUE, message=FALSE}
library(taxonbridge)
sample <- load_sample()
dim(sample)
```

Next, retrieve all rows that have lineage information in both the GBIF backbone and NCBI:

```{r, include=TRUE}
lineages <- get_lineages(sample)
```

Then validate the lineages by using the kingdom and family taxonomic ranks, and create a list of the resulting tibble(s). Note that phylum, class, and order may also be used. In this example, entries that failed validation are returned by setting `valid = FALSE`.

```{r, include= TRUE}
kingdom <- get_validity(lineages, rank = "kingdom", valid = FALSE)
family <- get_validity(lineages, rank = "family", valid = FALSE)
candidates <- list(kingdom, family)
```

Finally, detect candidate incongruencies (excluding those with uninomial scientific names):

```{r, include= TRUE}
get_inconsistencies(candidates, uninomials = FALSE)
```

Two binomial names exhibit incongruency. Upon reference to the literature and the individual entries it can be seen that: 

* *Attheya septentrionalis* is assigned to different families of the **problematica** order Chaetocerotales

* *Gordonia neofelifaecis* is a plant (family: Theaceae) in the GBIF but a bacterium in the NCBI (family: Gordoniaceae)

*Attheya septentrionalis* has the status "synonym" in the GBIF data:

```{r, include= TRUE}
lineages[lineages$canonicalName=="Attheya septentrionalis", "taxonomicStatus"]
```

Applying the `get_status()` function and rerunning the exercise leaves only *Gordonia neofelifaecis* as a binomial incongruency with biological provenance:

```{r, include= TRUE}
lineages <- get_status(get_lineages(sample), status = "accepted")
kingdom <- get_validity(lineages, rank = "kingdom", valid = FALSE)
family <- get_validity(lineages, rank = "family", valid = FALSE)
candidates <- list(kingdom, family)
get_inconsistencies(candidates, uninomials = FALSE)
```

## 2. Annotating a custom taxonomy

Again, start by loading the 2000 row sample dataset that comes with `taxonbridge`. Then apply the `get_taxa()` method to find all decapod crustaceans in the sample dataset: 

```{r, include=TRUE, message=FALSE}
library(taxonbridge)
sample <- load_sample()
decapoda <- get_taxa(sample, order = "decapoda")
```

The `decapoda` object will serve as your base taxonomy. Create a new object that only retains decapods known as swimming crabs:

```{r, include=TRUE, message=FALSE}
swimming_crabs <- get_taxa(sample, family = "portunidae")
```

Next annotate your base taxonomy with this colloquial name for the family Portunidae:

```{r, include=TRUE, message=FALSE}
decapoda <- annotate(decapoda, names = swimming_crabs$canonicalName, 
                     new_column = "swimming_crabs", present = "1")
```

A new column by the name "swimming_crabs" has been added to the base taxonomy:

```{r, include=TRUE, message=TRUE}
colnames(decapoda)
```

Since only the `present` parameter and not the `absent` parameter was passed to `annotate()`, all species that are not members of the Portunidae will be assigned `NA` by default in the `swimming_crabs` column. Swimming crabs can therefore be retrieved from the base taxonomy with the following command:

```{r, include=TRUE, message=TRUE}
decapoda[!is.na(decapoda$swimming_crabs),"canonicalName"]
```

## 3. Visualizing a custom taxonomy

Continue using the annotated base taxonomy from example 2. Prepare two distributions for the base taxonomy using `prepare_rank_dist()`:

```{r, include=TRUE, message=TRUE, fig.show="hold"}
GBIF_dist <- prepare_rank_dist(decapoda, GBIF = TRUE)
NCBI_dist <- prepare_rank_dist(decapoda, NCBI = TRUE)
plot_mdb(GBIF_dist)
plot_mdb(NCBI_dist)
```

The plots show that there is a difference between the entries of the NCBI and GBIF. Looking at the previously prepared distributions reveal that the NCBI lacks lineage data for two species:

```{r, include=TRUE, message=TRUE}
GBIF_dist
NCBI_dist 
```

Assuring that both the NCBI data and GBIF data have lineage data by using `get_lineages()` solves this problem at the cost of losing two GBIF entries that are not available in the NCBI:

```{r, include=TRUE, message=TRUE, fig.show="hold"}
lineages <- get_lineages(decapoda)
GBIF_dist <- prepare_rank_dist(lineages, GBIF = TRUE)
NCBI_dist <- prepare_rank_dist(lineages, NCBI = TRUE)
plot_mdb(GBIF_dist)
plot_mdb(NCBI_dist)
```

Note that `get_lineages()` should be used with care since extinct species in the GBIF are unlikely to have lineage data in the NCBI.

Now that both the GBIF and NCBI data have lineage information, the validity of the lineage information can be accessed in the same way as was done in example 1:

```{r, include=TRUE, message=TRUE}
get_validity(decapoda, valid = FALSE)
```

Two entries have invalid data:

* The species *Palibythus magnificus* belongs to the family Palinuridae in the GBIF but belongs to the family Synaxidae in the NCBI.

* The genus *Lysirude* belongs to the family Lyreididae in the GBIF but belongs to the family Raninidae in the NCBI.

Annotating the base taxonomy with these inconsistencies is a good idea:

```{r, include=TRUE, message=TRUE}
decapoda <- annotate(decapoda, get_validity(decapoda, valid = FALSE)$canonicalName,
                    new_column = "family_inconsistencies", present = 1)
colnames(decapoda)
decapoda[!is.na(decapoda$family_inconsistencies),"canonicalName"]
```
